---
title: "Excercise 6"
author: "Jussipekka Salo"
date: "2023-12-11"
output: html_document
---

# 6. Analysis of longitudinal data

```{r, echo=FALSE, warning=FALSE}
library(tidyverse)
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
RATSL <- readRDS("data/rats.rds")

```


```{r}

# Look at the (column) names of RATSL
names(RATSL)

# Look at the structure of RATSL
str(RATSL)

# Print out summaries of the variables
summary(RATSL)

```
Data contains information about the weight of rats in three different group. From each rat we have several obs in time. By looking at the structure, we can see that the data is in long format.

## Let's visualize the data, and see how the evolution of weight differs by group.
```{r}


#Access the package ggplot2
library(ggplot2)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  labs(title="Rat weight by group")+
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight))) +
  theme_minimal()

```


Looking at the graph it seems that there "tracking", meaning that rats who are heavier in the beginning of the diet are also heavy at the end of diet. Let's take a closer look of this phenomena, by standardizing the weight.


```{r}


library(dplyr)
library(tidyr)
# Standardise the variable bprs
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate( weight_std = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot again with the standardised bprs
library(ggplot2)
ggplot(RATSL, aes(x = Time, y = weight_std, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(limits = c(min(RATSL$weight_std), max(RATSL$weight_std))) +
  theme_minimal()

```

Clearly, the groups have level differences in terms of the weight.

## Next, let's use boxplots to visualize our data

```{r}
# Work with the exercise in this chunk, step-by-step. Fix the R code!
# BPRSL is available



# Summary data with mean and standard error of bprs by treatment and week 
RATSSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight) / sqrt(n())) %>% 
  ungroup()

# Glimpse the data
glimpse(RATSSS)

# Plot the mean profiles
ggplot(RATSSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(weight)")+
  theme_minimal()

```
The graph suggest that there are statistically significant differences in weight by group 1 and the others, but it is hard to say if there are differences between group 2 and 3, because the confidence intervals overlap.
Nevertheless, this graph does not tell if the diet is the reason we see the level differences, because the level differences were there already at day 1.

## Box-plots
```{r}
# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline TIME 1)
RATSL8S <- RATSL %>%
  filter(Time > 1) %>% # Drop the first week
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)

# Draw a boxplot of the mean versus treatment
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "Evolution of weight, Days 8-64")+
  theme_minimal()
```


Graph shows that there are outliers, so let us get rid off them.




```{r}
# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data

RATSL8S1 <- RATSL %>% 
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup() %>% 
  filter(!(mean < 250 & Group=="1")) %>% 
  filter(!(mean > 550 & Group =="2")) %>% 
  filter(!(mean < 500 & Group=="3"))

# Draw a boxplot of the mean versus treatment
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "Evolution of weight, Day 8-64")+
  theme_minimal()

```


Graph suggest that there are statistically significant differences between the weights of the groups.



## Anova test

```{r}


# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATSL8S %>% 
  mutate(baseline=RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```
We can't conclude that there are statistically significant differences between groups, since the p-value of F-test is bigger than the critical 0.05. In other words, we can not reject the (null) hypothesis that group means are equal to each other.


## Move on to the part 2. Let's use the BRLS data


```{r}
BPRSL <- readRDS("data/bprs.rds")

```


```{r}

# Look at the (column) names of BPRS
names(BPRSL)

# Look at the structure of BPRS
str(BPRSL)
dim (BPRSL)

# Print out summaries of the variables
summary(BPRSL)


```

Data contains info of 40 males who were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks.


```{r}
ggplot(BPRSL,aes(x=week,y=bprs,color=subject))+
  geom_line()+
  facet_grid(~treatment,labeller = label_both)+
  theme_minimal()
```


It seems that the average of bprs is very similar in both groups but there is an outlier in treatment 2.


## Linear model 
```{r}
BPRSL_reg <- lm(formula = bprs ~ week + treatment, data=BPRSL)
summary(BPRSL_reg)
```

Coefficient of treatment 2 is positive meaning that the score of bprs is higher in on average in treatment 2, but that is not statistically significant. Linear models assumes independence of obs, which is not realistic here because we have panel data. Let us use random effects.


## Random intercept model
```{r}

# access library lme4
library(lme4)

BPRSL_ref <- lmer( bprs ~ week + treatment + (1|subject) , data=BPRSL, REML = FALSE)

#Print the summary of the model
summary(BPRSL_ref)


```
Still the coefficient of treatment2 is not significant. Let see what happens when we will not only allow different intercepts for individuals but also different slopes.

```{r}
# create a random intercept and random slope model
library(lme4)

#Run the random intercept model
BPRSL_ref1 <- lmer( bprs ~ week + treatment + (week|subject) , data=BPRSL, REML = FALSE)

#Print the summary
summary(BPRSL_ref1)

#Anova test on the two models
anova(BPRSL_ref1,BPRSL_ref)
```

The coefficient of treatment2 is still not significant but by running the anova test, we can conclude that the fit is better when we include random intercept and slope. Let's try to do even better job by including interaction terms.


## Time to interact: Random Intercept and Random Slope Model with interaction

```{r}
#Run model with interactions
BPRSL_ref2 <- lmer( bprs ~ week + treatment + week*treatment + (week|subject) , data=BPRSL, REML = FALSE)

summary(BPRSL_ref2)


#Anova test on the interaction and without interactions models
anova(BPRSL_ref2,BPRSL_ref1)


```

Looks like anova gives p-value of 0.07 for chi-square test, meaning that there is no statistically significant differences between the models.

#Plot the fitted values


```{r}
# Create a vector of the fitted values
BPRSL$Fitted <- predict(object = BPRSL_ref2)

# draw the plot of RATSL with the Fitted values of weight
ggplot(BPRSL, aes(x = week, y = Fitted, color = subject)) +
  geom_line() +
  facet_grid(~treatment,labeller = label_both)+
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2)) +
  scale_y_continuous(name = "Fitted bprs") +
  theme(legend.position = "top")

```